---
title: "Emergence_Paper"
author: "Kriddie"
date: "2/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
#use here package to navigate to folder where your project is
library(here)
library(lubridate)
library(dplyr)
library(tidyr)
#install.packages("DataCombine")
library(DataCombine)
#note - MASS package may be masking functions needed for making pivot plots used by tidyr and dplyr
library(vegan)
library(MASS)

```

## about
This R markdown is to document manipulation and analyses of stonefly emergence data collected within the Flathead River watershed in 2017

#TO DO ID progress
* KW has Ided all samples until the 25 of July. about 10 vials left to ID - 2020-02-20
* still need help confirming ID


First, let's import the most current "picked list" and clean up the data

```{r import data}

df <- read.csv(here::here("Data_For_Analysis/Emergence_PickedList_2020-04-23.csv"))

#name Columns
colnames(df) <-  c("sampleID","SampleMethod","SampleMethodCode","year_collected","RiverCode","FloodplaneCode","SiteCode","TrapNo","SampleNo","CollectionDate","CollectionWeeknum","CollectionDay","Person.ID","IDDate","Person.IDConfirm","VialCount","Genus","Species","male.female.nymph","count")


df$SampleMethod <- as.factor(tolower(df$SampleMethod))
df$SampleMethodCode <- as.factor(toupper(df$SampleMethodCode))
df$year_collected <- as.integer(df$year_collected)
df$RiverCode <- as.factor(toupper(df$RiverCode))
df$FloodplaneCode <- as.factor(toupper(df$FloodplaneCode))
df$SiteCode <- as.factor(toupper(df$SiteCode))
df$SiteCode <- trimws(df$SiteCode, which = c("both", "left", "right"), whitespace = "[ \t\r\n]")
df$TrapNo <- as.integer(df$TrapNo)
df$SampleNo <- as.integer(df$SampleNo)
df$CollectionDate <- format(strptime(df$CollectionDate, format = "%d-%b-%y"), "%Y-%m-%d")
df$CollectionWeeknum <- as.integer(df$CollectionWeeknum)
df$CollectionDay <- as.integer(df$CollectionDay)
df$IDDate <- format(strptime(df$IDDate, format = "%d-%b-%y"), "%Y-%m-%d")
df$Genus <- as.factor(tolower(df$Genus))
df$Species <- as.factor(toupper(df$Species))
df$male.female.nymph <- as.factor(tolower(df$male.female.nymph))
df$count <- as.integer(df$count)

#add column combinig genus and species
df$genus_species <- NA
df$genus_species <- paste(df$Genus,df$Species,sep="_")
df$genus_species  <- gsub(" ", "", df$genus_species )


```
## To DO - Data Cleaning
More data cleaning needed
*check that sampleID matches with other column entry*
*Look through data to make sure sampleIDs and date collected make sense*
*missing dates -> look in field note books*
*missing species/genus -> look in previous versions of picked list*

## Explore the data by creating pivot plots ##

pivot plots will also be used to run NMDS


```{r pivot, echo=FALSE}

#get rid of SPP and blanks 
df <- df %>% mutate_all(na_if,"")
df <- df %>% filter(Species != "SPP"| NA)

df2 <- df[,c("RiverCode","FloodplaneCode","SiteCode", "genus_species","count","CollectionDate")]
#Create pivot plot that summerizes by select variables
df2 <- df2 %>% filter(CollectionDate < "2017-08-01")

pivot <- df2 %>%
  group_by(RiverCode, FloodplaneCode,SiteCode,genus_species)%>%        #select wich variable you want to 
  summarise(TotalCount = sum(count))                                    #select how you want to summerize

#Use pivot plot to create matrix that can be read by vegan package
pivot2 <- pivot %>%
  arrange(RiverCode) %>%
  spread(genus_species,TotalCount) %>%
  filter(SiteCode != 	"UNKNOWN2") %>%
  filter(FloodplaneCode != "PT") %>%
  filter(SiteCode != "PO") %>%
  filter(SiteCode != "DT") %>%
  filter(SiteCode != "BC") %>%
  filter(SiteCode != "PB")
  
  
  

#remove unnecessary variables (we may need to combine variables here to make a more descriptive column - ie weeknum.by.site)
pivot2$SiteCode = paste(pivot2$FloodplaneCode , pivot2$SiteCode, sep="_")
pivot2$FloodplaneCode <- NULL
pivot2$RiverCode <- NULL
pivot2$CollectionDate <- NULL

#format for vegan package
pivot2 <-as.data.frame(pivot2)

add_row_names <- pivot2$SiteCode
pivot2$SiteCode <- NULL
row.names(pivot2) <- add_row_names

#species na
stoneflyDF <- pivot2
stoneflyDF[is.na(stoneflyDF)] = 0


```




**NMDS Plot**
An NMDS plot places each site on an X Y coordinate plane. The closer the sites, the more similar the species composition.
Link to helpful tutorial for nmds using vegan:
*https://jonlefcheck.net/2012/10/24/nmds-tutorial-in-r/*


The following two plots are stress plots and NMDS plots respectively. The following NMDS plot compares species composition at sites by combining all species collected at each site

###NMDS Plot 1
```{r, message=FALSE, warning=FALSE, nmds, echo=FALSE}


# Calculating relative abundance and creating new dataframe with relative abundance data
stoneflyDF.spp.rel <-         
  decostand(stoneflyDF, method = "total")

# Calculate distance matrix
stoneflyDF.spp._distmat <- 
  vegdist(stoneflyDF.spp.rel, method = "bray")

# Creating easy to view matrix and writing .csv
stoneflyDF.spp._distmat <- 
  as.matrix(stoneflyDF.spp._distmat, labels = T)
#write.csv(stoneflyDF.spp._distmat, "stoneflyDF.spp._distmat.csv")

# Running NMDS in vegan (metaMDS)
stoneflyDF.spp_NMS <-
  metaMDS(stoneflyDF.spp._distmat,
          distance = "bray",
          k = 2,
          maxit = 999, 
          trymax = 500,
          wascores = TRUE)


# Shepards test/goodness of fit
goodness(stoneflyDF.spp_NMS) # Produces a results of test statistics for goodness of fit for each point

stressplot(stoneflyDF.spp_NMS) # Produces a Shepards diagram

plot(stoneflyDF.spp_NMS, "sites")# Produces distance 

#add in river code
treat=c(rep("KA",5),rep("MF",5),rep("NF",7))
#new plot
ordiplot(stoneflyDF.spp_NMS,type="n")
#draw polygon around similar sites
ordihull(stoneflyDF.spp_NMS,groups=treat,draw="polygon",col="grey90",label=F)
orditorp(stoneflyDF.spp_NMS,display="sites",col=c(rep("green",5),rep("blue",5),rep("red",7)),
         air=0.01,cex=1.25)

```

this NMDS plot is based on a matrix of data without calculating relative abundance or distance.
I'm not sure wich is more appropriate. It is nice that we can also display species names in the one below but it is very messy

###NMDS Plot 2
```{r test.NMDS, echo=FALSE}


#plot with names

test.NMDS=metaMDS(stoneflyDF, # Our community-by-species matrix
                  k=2) # The number of reduced dimensions
stressplot(test.NMDS)
plot(test.NMDS)

ordiplot(test.NMDS,type="n")
orditorp(test.NMDS,display="sites",cex=1.25,air=0.01)
orditorp(test.NMDS,display="species",col="red",air=0.01)
 

#plot with "treatment shapes"

treat=c(rep("KA",5),rep("MF",5),rep("NF",7))
ordiplot(test.NMDS,type="n")
ordihull(test.NMDS,groups=treat,draw="polygon",col="grey90",label=F)
orditorp(test.NMDS,display="species",col="grey20",air=0.01)
orditorp(test.NMDS,display="sites",col=c(rep("green",5),rep("blue",5),rep("red",7)),
         air=0.01,cex=1.25)



```
