---
title: "Emergence_Paper"
author: "Kriddie"
date: "2/18/2020"
output: html_document
---

```{r setup, include=FALSE}
knitr::opts_chunk$set(echo = TRUE)
```

## about
This R markdown is to document manipulation and analyses of stonefly emergence data collected within the Flathead River watershed in 2017

#TO DO ID progress
* KW has Ided all samples until the 25 of July. about 10 vials left to ID - 2020-02-20
* still need help confirming ID


First, let's import the most current "picked list" and clean up the data

```{r import data}
#use here package to navigate to folder where your project is
library(here)
library(lubridate)
library(dplyr)
library(tidyr)
#df <- read.csv(here::here("/Emergence_PickedList_2020-02-18.csv"))
df <- read.csv(here::here("Data_For_Analysis/Emergence_PickedList_2020-04-23.csv"))

#name Columns
colnames(df) <-  c("sampleID","SampleMethod","SampleMethodCode","year_collected","RiverCode","FloodplaneCode","SiteCode","TrapNo","SampleNo","CollectionDate","CollectionWeeknum","CollectionDay","Person.ID","IDDate","Person.IDConfirm","VialCount","Genus","Species","male.female.nymph","count")


df$SampleMethod <- as.factor(tolower(df$SampleMethod))
df$SampleMethodCode <- as.factor(toupper(df$SampleMethodCode))
df$year_collected <- as.integer(df$year_collected)
df$RiverCode <- as.factor(toupper(df$RiverCode))
df$FloodplaneCode <- as.factor(toupper(df$FloodplaneCode))
df$SiteCode <- as.factor(toupper(df$SiteCode))
df$TrapNo <- as.integer(df$TrapNo)
df$SampleNo <- as.integer(df$SampleNo)
df$CollectionDate <- format(strptime(df$CollectionDate, format = "%d-%b-%y"), "%Y-%m-%d")
df$CollectionWeeknum <- as.integer(df$CollectionWeeknum)
df$CollectionDay <- as.integer(df$CollectionDay)
df$IDDate <- format(strptime(df$IDDate, format = "%d-%b-%y"), "%Y-%m-%d")
df$Genus <- as.factor(tolower(df$Genus))
df$Species <- as.factor(toupper(df$Species))
df$male.female.nymph <- as.factor(tolower(df$male.female.nymph))
df$count <- as.integer(df$count)




```
## To DO - Data Cleaning
More data cleaning needed
*check that sampleID matches with other column entry*
*Look through data to make sure sampleIDs and date collected make sense*
*missing dates -> look in field note books*
*missing species/genus -> look in previous versions of picked list*

## Explore the data by creating pivot plots ##

pivot plots will also be used to run NMDS


```{r pivot, echo=FALSE}

#Create pivot plot that summerizes by select variables

#MASS package maskes "select" in dplyr. So I specify dplyr select below

pivot <- df %>%
  dplyr::select(FloodplaneCode,SiteCode, Species,count) %>%  #select variabls
  group_by(FloodplaneCode,SiteCode,Species)%>%        #select wich variable you want to group by
  summarise(TotalCount = sum(count))                                    #select how you want to summerize

#Use pivot plot to create matrix that can be read by vegan package
pivot2 <- pivot %>%
  spread(Species,TotalCount)

#remove unnecessary variables (we may need to combine variables here to make a more descriptive column - ie weeknum.by.site)
pivot2 <-as.data.frame(pivot2)
pivot2[,1] <- NULL
#Use first row (key variable) as column name
rownames(pivot2) <- pivot2[,1]
pivot2[,1] <- NULL

#get rid of SPP and blanks 
pivot2 <- dplyr::select(pivot2, -c(SPP, V1))


```




**NMDS Plot**
An NMDS plot places each site on an X Y coordinate plane. The closer the sites, the more similar the species composition.
Link to helpful tutorial for nmds using vegan:
*https://jonlefcheck.net/2012/10/24/nmds-tutorial-in-r/*


The following two plots are stress plots and NMDS plots respectively. The following NMDS plot compares species composition at sites by combining all species collected at each site

###NMDS Plot 1
```{r, message=FALSE, warning=FALSE, nmds, echo=FALSE}

#note - vegan package may be masking functions needed for making pivot plots used by tidyr and dplyr

library(vegan)
library(MASS)

#replace all NA with zeros
pivot2[is.na(pivot2)] <- 0


#sort data
#data.by.river <- test.data2[order(test.data2$river, test.data2$floodplane),]

#exploring the data
#abundance.matrix <- data.by.river[,4:54]
csum <- colSums(pivot2)
any(is.na(csum))
which(is.na(csum))


test.NMDS=metaMDS(pivot2, # Our community-by-species matrix
                  k=2) # The number of reduced dimensions
stressplot(test.NMDS)
plot(test.NMDS)


```



These next plot adds color and labels for easier plot interpretation

###NMDS Plot 2
```{r test.NMDS, echo=FALSE}
#plot with names

ordiplot(test.NMDS,type="n")
orditorp(test.NMDS,display="species",col="red",air=0.01)
orditorp(test.NMDS,display="sites",cex=1.25,air=0.01)

#plot with "treatment shapes"

treat=c(rep("MF",8),rep("KA",5),rep("NF",7))
ordiplot(test.NMDS,type="n")
#ordihull(test.NMDS,groups=treat,draw="polygon",col="grey90",label=F)
orditorp(test.NMDS,display="species",col="grey20",air=0.01)
orditorp(test.NMDS,display="sites",col=c(rep("green",8),rep("blue",5),rep("red",7)),
         air=0.01,cex=1.25)

```

This plot shows how community composition relates to site elevation

###NMDS Plot 3
```{r environmental gradient, echo=FALSE}
#plot with environmental gradient
elevation=data.by.river[,3]
ordisurf(test.NMDS,elevation,main="",col="forestgreen")
# Finally, display species on plot
orditorp(test.NMDS,display="species",col="grey30",air=0.1,
         cex=1)
```